---
title: "Celiac Disease Polygenic Risk Score Based on Sharp et al."
author: "Lucas Gillenwater"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---
# Background
Sharp et al. created a polygenic risk score (GRS) for Celiac Disease using the results from several large GWAS studies. The GRS multiplies the observed dosage for the risk allele by the logOdds for each locus. When repeated in the HTP cohort, the model results in a lower ROC than was observed by Sharp et al. This analysis explores how to optimize the model to improve this ROC. 

# Recreating the GRS
In this analysis I attempted to recreate the GRS from the Sharp et al. study. An issue with the smaller HTP data is the case imbalance. There are 19 observed Celiac cases with 185 controls. This imbalance can cause bias in the models from outliers in the small sample of cases. To overcome this bias I repeatedly subsampled from the data, preserving the case/control ratio. At each iteation I calculated the logOdds and after 1000 iterations chose the best performing model to be included in the score. I then calculated the AUC for the GRS using the HTP weights and compared that to the AUC using the Sharp et al. weights (using the HTP dosages as input). 




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# load in the required libraries
require(tidyverse)
require(precrec)
require(caret)
require(ggplot2)
require(pROC)
library(data.table)
library(spatstat)
# ste the seed for reproducibility
set.seed(1234)
```


```{r load and filter data}
# Data
meta  <- read.csv("./data/MEGA_041822_META_CeliacGRS_v0.1_JRS.csv")# Load the metadata
results  <- read.csv('./data/MEGA_041822_RESULTS_CDGRS2022_VariantDosage_vs_Celiac_v0.2_JRS.csv')# Load the Results
analysis  <- read.csv('./data/MEGA_041822_AnalysisData_CDGRS_Sharp2022_v0.1_JRS.csv')# Load the Analysis data
pcs <- read.csv('./data/MEGA_041822_Espinosa_MEGA2_HTP_GS_08132019_updated_callrate_passing_QC_KEEPforHLAvsCeliac_EXCLUDEvariants_mind05_geno0.02_maf0.05_PRUNEDindeppairwise0.2_v0.1_JRS.eigenvec', sep = " ")


# Filtering #
tokeep  <- meta %>% # create a vector of ID's to keep. Some need to be removed based on relatedness.
    filter(EXCLUDE_from_analysis == 0)%>%
    .$MEGA.IID

analysis_data  <- analysis %>% # Filter the data
    filter(MEGA.IID %in% tokeep)

# Create dataset for producing the genetic risk score
Sharp_risk_score <- analysis_data %>%
  select('MEGA.IID', 'GRS') %>% 
  distinct(MEGA.IID, .keep_all = T) %>% 
  left_join(meta[, c("MEGA.IID", "Celiac")], by = "MEGA.IID")

# Create DS dosage data frame
DS_dosage <- analysis_data %>% 
  select(MEGA.IID, Variant..Sharp.2022.,Dosage_for_GRS) %>% 
  pivot_wider(id_cols = 'MEGA.IID',
              names_from = 'Variant..Sharp.2022.',
              values_from = 'Dosage_for_GRS') %>%
    left_join(meta[, c('MEGA.IID', 'Celiac')], by = 'MEGA.IID') %>%
    left_join(pcs[, c('IID', 'PC1', 'PC2','PC3','PC4','PC5')], by = c('MEGA.IID'= 'IID')) %>%
    column_to_rownames('MEGA.IID')


celiac_subjects <- which(dosage_data$Celiac == 1)
control_subjects <- seq(1:204)[-celiac_subjects]

```

```{r logistic regression model} 

celiac_log_reg <- function(subjects = subjects, DS_dosage = DS_dosage, control_subjects = control_subjects, celiac_subjects = celiac_subjects, control_to_case_ratio = 3, snp = snp ) {

    train_samples  <-sample(length(control_subjects), (length(celiac_subjects)-length(subjects))*control_to_case_ratio)
    train_controls <- control_subjects[train_samples]
    tmp_data <- DS_dosage[c(celiac_subjects[-subjects],train_controls),]
    tmp <- as.formula(paste0('Celiac ~ ',snp,' + PC1 + PC2+PC3+PC4+PC5'))
    tmp_model <- suppressWarnings(glm(tmp,family= binomial(link = "logit"),data = tmp_data))
    weight <- coef(tmp_model)[2]
    weight <- ifelse(is.finite(weight), weight, NA)
    tmp_resutls <- c(weight,  suppressMessages(confint(tmp_model)[2,]),  tmp_model$aic, ifelse(is.na(weight), NA, summary(tmp_model)$coefficients[2,4]))
    return(list(logreg = tmp_model,train_samples = train_samples))
}

```

```{r accuracy}

celiac_accuracy <- function(subjects = subjects, DS_dosage = DS_dosage, control_subjects = control_subjects, celiac_subjects = celiac_subjects, test_iterations = 10, control_to_case_ratio = 3, snp = snp) {

    model <- celiac_log_reg(subjects=subjects, DS_dosage=DS_dosage, control_subjects=control_subjects, celiac_subjects=celiac_subjects, control_to_case_ratio= control_to_case_ratio, snp=snp)
    
 #weight = model$coef[2]
    
    tmp_accuracy <- list()
    for(x in 1:test_iterations){
        test_controls <- sample(control_subjects[-model$train_samples],control_to_case_ratio)
        test_data <- DS_dosage[c(subjects, test_controls),c(snp,'Celiac', 'PC1', 'PC2','PC3','PC4','PC5')]
        probabilities <- model$logreg %>% predict(test_data, type = "response")
        test_data$prediction <- ifelse(probabilities > (1/(case_to_control_ratio+length(subjects))), 1, 0) # weighting predictions based on the ratio between cases and controls
        tmp_accuracy[[x]] <- mean(test_data$prediction == test_data$Celiac)
       }

    return(list(model = model, accuracy = median(as.numeric(tmp_accuracy))))
}
```
```{r control_sampling}
celiac_control_sampling <- function(subjects= subjects, DS_dosage = DS_dosage, control_subjects = control_subjects, celiac_subjects = celiac_subjects, test_iterations = 10, control_to_case_ratio = 3, snp = snp, sampling_iterations = 10) {


    all_models <- list()
    for(i in 1:sampling_iterations) {
        all_models[[i]] <- celiac_accuracy(subjects=subjects, DS_dosage=DS_dosage, control_subjects=control_subjects, celiac_subjects=celiac_subjects, test_iterations = test_iterations, control_to_case_ratio=control_to_case_ratio, snp = snp)
        
    }

    weights <- sapply(all_models, function(x) x$model$logreg$coefficients[2])
    accuracies <- sapply(all_models, function(x) x$accuracy)
    weighted_median_weight <- weighted.median(weights, accuracies)
    return(list(weighted_median_weight = weighted_median_weight, median_accuracy = median(accuracies)))
    
  
}
```

```{r cv}

celiac_cv  <-  function( DS_dosage = DS_dosage, control_subjects = control_subjects, celiac_subjects = celiac_subjects, test_iterations = 10, control_to_case_ratio = 3, snp = snp, sampling_iterations = 10, k = 1) {
    group_numbers <- ceiling(length(celiac_subjects)/k)
    groups <- split(celiac_subjects, cut(seq_along(celiac_subjects), group_numbers, labels = FALSE))
    weighted_medians <- list()
    for(i in 1:length(groups)) {
        weighted_medians[[i]] <- celiac_control_sampling(groups[[i]], DS_dosage=DS_dosage, control_subjects = control_subjects, celiac_subjects= celiac_subjects, control_to_case_ratio=control_to_case_ratio, snp=snp, sampling_iterations = sampling_iterations, test_iterations = test_iterations)
    }

    weights <- sapply(weighted_medians, function(x) x$weighted_median_weight)
    accuracies <- sapply(weighted_medians, function(x) x$median_accuracy)
    weighted_median_weight <- weighted.median(weights, accuracies)
    return(list(weighted_median_weight = weighted_median_weight, median_accuracy = median(accuracies), accuracies = accuracies))
}
        

```


```{r risk score cv function}

celiac_GRS <- function(DS_dosage = DS_dosage, control_subjects = control_subjects, celiac_subjects = celiac_subjects, test_iterations = 10, control_to_case_ratio = 3, snp = snp, sampling_iterations = 10, k = 1) {

    original_names <- names(DS_dosage)
    names(DS_dosage) <- make.names(original_names)

    options(warn = -1)# suppress the warning about fitting values to 0 or 1
    # identify subjects
   
  #create a list for results
    grs_results <- list()
# 50 test_iteations, 50 sampling_iterations
  for(snp in names(DS_dosage)[1:49]){
    print(snp)
    # test for differences in Celiac's disease by dosage

    breakdown <- table(DS_dosage[[snp]], dosage_data$Celiac)
    
    if(!(dim(breakdown)[1] == 2 && any(breakdown == 0))) {
        grs_results[[snp]] <- celiac_cv(DS_dosage = DS_dosage, control_subjects = control_subjects, celiac_subjects = celiac_subjects, test_iterations = test_iterations, control_to_case_ratio = control_to_case_ratio, snp = snp, sampling_iterations = sampling_iterations, k = k)
        print(paste0(round(which(names(DS_dosage)==snp)/49,2)*100,"% complete"))


    } else {
        grs_results[[snp]] = list(weighted_median_weight = NA, median_accuracy = NA, accuracies = NA )
        print(paste0(round(which(names(DS_dosage)==snp)/49,2)*100,"% complete"))

    }
  }
    return(grs_results)
}


```
 
 
 
 
 
```{r}

grs_wrapper <- function(DS_dosage = DS_dosage, control_subjects = control_subjects, celiac_subjects = celiac_subjects, test_iterations = 10, control_to_case_ratio = 3, snp = snp, sampling_iterations = 10, k = 1) {
                                        # reset warning settings 
    options(warn = 0)

    grs_results <- celiac_GRS(DS_dosage = DS_dosage, control_subjects = control_subjects, celiac_subjects = celiac_subjects, test_iterations = test_iterations, control_to_case_ratio = control_to_case_ratio, snp = snp, sampling_iterations =sampling_iterations, k = k)
    
    grs_results_df <- data.frame(do.call(rbind,grs_results))
    rownames(grs_results_df) <- names(DS_dosage)[1:49]
  
                                        # create GRS
    GRS <- colSums(apply(DS_dosage[,1:49], 1, function(x) x* as.numeric(grs_results_df$weighted_median_weight)), na.rm = TRUE)

  # plot the roc curves
    rocs <- list()
    rocs[["HTP"]] <-  suppressMessages( pROC::roc(response = as.numeric(DS_dosage$Celiac), predictor= as.numeric(GRS)))
    rocs[["Sharp"]] <- suppressMessages(pROC::roc(Sharp_risk_score$Celiac, Sharp_risk_score$GRS))

    p1 <- ggroc(rocs) +
        theme_classic()
    sharp_scores <- analysis %>% 
        select(Score_Weight..logOR., Variant..Sharp.2022.) %>% 
        distinct(.keep_all = TRUE)
    

    scores <- cbind(sharp_scores, as.numeric(grs_results_df$weighted_median_weight))
    names(scores) <- c("Sharp", "Variant", "HTP")
    scores <- scores %>% 
        pivot_longer(cols = c(Sharp, HTP))
    
    
    
    p2 <- ggplot(scores, aes(y= value, x=Variant, fill = name)) +
        geom_bar(stat = "identity", alpha = .6, position = "dodge") +
        theme_classic() +
        theme(axis.text.x = element_text(angle = 90))
    
                                        # pdf("HTP_Sharp_weights_072722.pdf")
                                        # p2
                                        # dev.off()
    accuracies <- data.frame(do.call(rbind, grs_results_df$accuracies))
    accuracies <- accuracies %>%
            rownames_to_column() %>%
        pivot_longer( !rowname)
        
    
    p3 <- ggplot(accuracies, aes(x = rowname, y = value, fill = rowname)) +
        geom_boxplot() +
        coord_flip() +
        xlab("accuracy")+
        ylab("Locus") +
        theme_classic() +
        theme(legend.position = "none")


    print(rocs)
    return(list(GRS = GRS, ROC_plot = p1, weights_plot = p2, accuracies = p3))
  
}


```



```{r application}

k1 <- grs_wrapper(DS_dosage = DS_dosage, control_subjects = control_subjects, celiac_subjects = celiac_subjects, test_iterations = 3 , control_to_case_ratio = 3, snp = snp, sampling_iterations = 3, k = 1)
k2 <- grs_wrapper(DS_dosage = DS_dosage, control_subjects = control_subjects, celiac_subjects = celiac_subjects, test_iterations = 10, control_to_case_ratio = 3, snp = snp, sampling_iterations = 10, k = 2)
k3 <- grs_wrapper(DS_dosage = DS_dosage, control_subjects = control_subjects, celiac_subjects = celiac_subjects, test_iterations = 5, control_to_case_ratio = 3, snp = snp, sampling_iterations = 5, k = 4)
```



```{r sub sampling test, eval = FALSE}

sub18 <- htpGRS(pseudoCount = 0, iterations = 10, subN = 18)
## sub17 <- htpGRS(pseudoCount = 0, iterations = 1000, subN = 17)
## sub16 <- htpGRS(pseudoCount = 0, iterations = 1000, subN = 16)
## sub15 <- htpGRS(pseudoCount = 0, iterations = 1000, subN = 15)
## sub14 <- htpGRS(pseudoCount = 0, iterations = 1000, subN = 14)

## sub_results = list(sub18 = sub18, sub17 = sub17, sub16 = sub16, sub15 = sub15, sub14 = sub14)
## save(sub_results, file = "sub_sampling_results.Rdata")

# Using 18 samples each iterations results in the best performing model. 

```

# Sub sampling sensitivity test

To find the correct number of cases to sub sample to, I calculated the GRS for different numbers of sampled cases (14-18). The highest AUC of 0.867 was observed for sub sampling with 18 cases. 

```{r sub sampling plot, message=FALSE}
load("sub_sampling_results.Rdata")

sub_aucs <- sapply(1:5, function(x) roc(response = DS_dosage$Celiac, predictor = sub_results[[x]]$GRS)$auc)

plot(18:14, sub_aucs, xlab = "number of cases", ylab = "AUC", type = "b")

```

# ROC plot

```{r}
sub_results$sub18$ROC_plot
```

# Weights
```{r}
sub_results$sub18$weights_plot
```

